---
title: "BRFSS Analysis (Guam 2020 dataset)"
author: "Tristan_Paulino"
date: "Last updated: 2023-12-10"
output:
  html_document:
    df_print: paged
  prettydoc::html_pretty:
    theme: journal
---

# Introduction

This documentation is supplementary to the manuscript draft "<i>The Epidemiology of Gout on Guam and the impact of areca nut</i>" which replicates the analysis, results, and displays the respective statistical tables. 

Our main objectives were to:

1. Estimate the current prevalence of gout in Guam utilizing the 2020 BRFSS data.
2. Describe the relationship between areca nut and gout.

### Step 1: Calling the libraries

Note that I am using the `gtsummary` package to create nice and near publication-ready tables/regression outputs. This package can handle generating simple frequencies and automate regression/ANOVA tables.

For all tables, I have reported

1. Unweighted sample size (N)
2. Unweighted sample (n)
3. Weighted percent/proportion (%)


```{r library, echo=TRUE, message=FALSE}
library(tidyverse)
library(haven)
library(survey) 
library(gtsummary)
library(forcats)

```
### Step 2: Importing SPSS dataset file into R using the haven package 


```{r data_upload, echo=TRUE, message=FALSE}

# Read in provided spss file into R using haven package
spss_data <- read_sav("GU20FINLvSplit_V3.SAV")   

# BRFSS Variabales in crude dataset
# v3 == Household Density Stratum Code
# v4 == Primary Sampling Unit
# v75 == Calculated Final Weight
# v140 == Age stratified in 6 groups
# SEXVAR == Respondent Sex
# Ethnic1 == Ethnicity
# v145 == Education Level Category
# GoutDiag == Age of Gout Diagnosis
# GoutMeds == Gout Medication Use
# ALCDAY5 == Alcohol Use Last 30 days
# SugarBev1 & SugarBev2 == consumption of fruit drinks and soda (two variables)
# BetChw2 == current areca/betel nut use 
# v148 == current smoking tobacco status
# USENOW3 == current smokeless tobacco use
# DIABETE4 == Diabetes diagnosis
# CVDCRHD4 == Cardiovascular disease diagnosis
# v142 == Obesity Status

guam_data <- spss_data %>% 
  select(v3,
         v4,
         v75,
         SEXVAR,
         v145,
         Ethnic1,
         v140,
         GoutDiag,
         GoutMeds,
         ALCDAY5,
         SugarBev1,
         SugarBev2,
         BetChw2,
         SMOKDAY2,
         USENOW3,
         DIABETE4,
         CVDCRHD4,
         v142)
```

### Step 3: Data cleaning prior to analyses

```{r data_wrangling, echo=TRUE, message=FALSE}
# Data Cleaning & Re-coding 
# Any unknown or missing value in a variable will be treated as "NA" and be excluded from analyses.

guam_data <- guam_data %>%
  mutate(
    edu_status = as_factor(case_when(
      v145 %in% c(1, 2) ~ "High School or below",
      v145 == 3 ~ "Some College",
      v145 == 4 ~ "College Graduate or higher",
      v145 == 9 ~ NA,
      TRUE ~ as.character(v145)
    )),
    sex = as_factor(case_when(
      SEXVAR == 1 ~ "Male",
      SEXVAR == 2 ~ "Female",
      TRUE ~ as.character(SEXVAR)
    )),
    ethnicity = as_factor(case_when(
      Ethnic1 %in% c(1, 15) ~ "CHamoru",
      Ethnic1 == 2 ~ "Filipino",
      Ethnic1 %in% c(9:14) ~ "Other Micronesian",
      Ethnic1 %in% c(3, 4, 5, 6, 7, 8, 16, 17) ~ "Other",
      Ethnic1 %in% c(26, 77, 99) ~ NA,
      TRUE ~ as.character(Ethnic1)
    )),
    age = as_factor(case_when(
      v140 == 1 ~ "18 to 24",
      v140 == 2 ~ "25 to 34",
      v140 == 3 ~ "35 to 44",
      v140 == 4 ~ "45 to 54",
      v140 == 5 ~ "55 to 64",
      v140 == 6 ~ "55 to 64",
      TRUE ~ as.character(v140)
    )),
    gout_status = as_factor(case_when(
      GoutDiag <= 95 ~ "Yes",
      GoutDiag == 96 ~ "No",
      GoutDiag %in% c(97,99) ~ NA,
      TRUE ~ as.character(GoutDiag)
    )),
    gout_med = as_factor(case_when(
      GoutMeds == 1 ~ "Never",
      GoutMeds %in% c(2,3) ~ "Sometimes",
      GoutMeds == 4 ~ "Regularly",
      GoutMeds == 5 ~ "Regularly",
      GoutMeds %in% c(7,9) ~ NA,
      TRUE ~ as.character(GoutMeds)
    )),
    alcohol = as_factor(case_when(
      ALCDAY5 %in% c(777,999) ~ NA,
      ALCDAY5 == 888 ~ "No",
      ALCDAY5 >= 101 & ALCDAY5 <= 230 ~ "Yes"
    )),
    sugar_bev = as_factor(case_when(
      SugarBev1 %in% c(777, 999) | SugarBev2 %in% c(777, 999) ~ NA,
      SugarBev1 == 888 | SugarBev2 == 888 ~ "No",
      TRUE ~ "Yes"
    )),
    betel_nut = as_factor(case_when(
      BetChw2 %in% c(1,2) ~ "Yes",
      BetChw2 == 3 ~ "No",
      BetChw2 %in% c(7,9) ~ "No"
    )),
    smoking = as_factor(case_when(
      SMOKDAY2 %in% c(1,2) ~ "Yes",
      SMOKDAY2 == 3 ~ "No"
    )),
    smokless = as_factor(case_when(
      USENOW3 %in% c(1,2) ~ "Yes",
      USENOW3 == 3 ~ "No",
      USENOW3 == 7 ~ NA
    )),
    diabetes = as_factor(case_when(
      DIABETE4 %in% c(1,2) ~ "Yes",
      DIABETE4 %in% c(3,4) ~ "No",
      DIABETE4 == 7 ~ NA
    )),
    heart_disease = as_factor(case_when(
      CVDCRHD4 == 1 ~ "Yes",
      CVDCRHD4 == 2 ~ "No",
      CVDCRHD4 == 7 ~ NA,
    )),
    obesity = as_factor(case_when(
      v142 %in% c(1,2,3) ~ "No",
      v142 == 4 ~ "Yes"
    ))
  )

```

### Step 4: Table 1.0 -- Demographic Description

```{r table1, echo=TRUE, message=FALSE}
# Creates the dataframe of demographic variables of interest plus BRFSS weights
table1_df <- guam_data %>% 
  select(
    v3,
    v4,
    v75,
    age,
    sex,
    ethnicity,
    edu_status
  )

# Creates survey object, specifies variables vs. weight
table1_survey <- svydesign(
  ids = ~v4,
  strata = ~v3,
  weights = ~v75,
  data = table1_df
)

# Creates Frequency Table of variables of interest
table1 <- tbl_svysummary(table1_survey,
               by = NULL,
               digits = list(
                 age ~ c(0,1),
                 sex ~ c(0,1),
                 ethnicity ~ c(0,1),
                 edu_status ~ c(0,1)
               ),
               include = c(age,sex,ethnicity,edu_status),    
               label = list(
                 age ~ "Age",
                 sex ~ "Sex",
                 ethnicity ~ "Ethnicity",
                 edu_status ~ "Education"),
               statistic = list(
                 all_categorical() ~ "{n_unweighted} ({p}%)"),
               missing = "no"
) %>% 
  modify_header(
    update = list(
      stat_0 ~ "**Overall** N = {n_unweighted}")
  ) %>% 
  as_gt() %>%
  gt::tab_header("Table 1: Demographic Description, Guam BRFSS 2020") %>% 
  gt::tab_options(
    table.font.size = "small",
    data_row.padding = gt::px(1)
  )

```

Here is what table 1 looks like. 

```{r table1out, echo=TRUE, results='asis'}
print(table1)
```

### Step 5: Table 2.0 Gout Prevalence & Gout Medication Use

```{r table2, echo=TRUE, message=FALSE}

table2_df <- guam_data %>% 
  select(
    v3,
    v4,
    v75,
    gout_status,
    gout_med
  )

table2_survey <- svydesign(
  ids = ~v4,
  strata = ~v3,
  weights = ~v75,
  data = table2_df
)

table2 <- tbl_svysummary(table2_survey,
                         by = NULL,
                         include = c(gout_status,gout_med),  
                         label = list(
                           gout_status ~ "Gout",
                           gout_med ~ "Gout Medication Use"
                           ),
                         statistic = list(
                           all_categorical() ~ "{n_unweighted} ({p}%)"),
                         missing = "no"
) %>% 
  modify_header(
    update = list(
      stat_0 ~ "**Overall** N = {n_unweighted}")
  ) %>% 
  as_gt() %>%
  gt::tab_header("Table 2: Gout Status & Medicine Use, Guam BRFSS 2020") %>% 
  gt::tab_options(
    table.font.size = "small",
    data_row.padding = gt::px(1)
  )

```

Here is what table 2 looks like. 

```{r table2out, echo=TRUE, results='asis'}
print(table2)
```

### Step 6: Table 3.0 Variables among Gout Participants.

```{r table3, echo=TRUE, message=FALSE}

# Creates the dataframe of variables of interest plus BRFSS weights
table3_df <- guam_data %>% 
  select(
    v3,
    v4,
    v75,
    gout_status,
    gout_med,
    ethnicity,
    sex,
    alcohol,
    betel_nut,
    smoking,
    sugar_bev,
    heart_disease,
    diabetes,
    obesity
  )

# Creates survey object, specifies variables vs. weight
table3_survey <- svydesign(
  ids = ~v4,
  strata = ~v3,
  weights = ~v75,
  data = table3_df
)  

# Creates Table of variables of interest
table3 <- tbl_svysummary(table3_survey,
               by = gout_status,
               missing = "no",
               include = c(ethnicity,sex,alcohol,betel_nut,smoking,sugar_bev,heart_disease,diabetes,obesity),
               label = list(ethnicity ~ "Ethnicity",
                            sex ~ "Sex",
                            alcohol ~ "Alcohol Consumption",
                            betel_nut ~ "Betel Nut Use",
                            smoking ~ "Smoking",
                            sugar_bev ~ "Sugar Beverage Consumption",
                            heart_disease ~ "Cardiovascular Disease",
                            diabetes ~ "Diabetes",
                            obesity ~ "Obesity"),
               statistic = list(
                 all_categorical() ~ "{n_unweighted} ({p}%)")
) %>%
  add_overall(
    col_label = "**Overall**, N {n_unweighted}"
  ) %>% 
  modify_header(
    update = list(
      label ~ "**Factors** ",
      stat_2 ~ "**Yes** N = {n_unweighted}",
      stat_1 ~ "**No** N = {n_unweighted}"
    )
  ) %>% 
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Gout Status**") %>% 
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>% 
  as_gt() %>%
  gt::tab_header("Table 3: Factors among people with Gout, Guam 2020 BRFSS") %>% 
  gt::tab_options(
    table.font.size = "small",
    data_row.padding = gt::px(1)
  )

```

Here is what table 3 looks like. 

```{r table3out, echo=TRUE, results='asis'}
print(table3)
```

### Step 7: Table 4.0 
Though I've done regressions in the past, it's not necessarily my forte. I conducted a logistic regression to see how well betel nut use can predict gout status. Based on this information, betel nut is not a good predictor for gout and not associated with it. This is also supported by the lack of significance observed in our chi-squre result (p value > 0.05)

```{r table4, echo=TRUE, message=FALSE}
# Creates the dataframe of betel nut use and gout variables plus BRFSS weights

# Convert betel_nut to a factor if it's not already
table4_df <- guam_data %>% 
  select(
    v3,
    v4,
    v75,
    gout_status,
    betel_nut
  )

table4_df$betel_nut <- relevel(table4_df$betel_nut, ref = "No")

# Creates the survey object needed for the logistic regression 
table4_survey <- svydesign(
  ids = ~v4,
  strata = ~v3,
  weights = ~v75,
  data = table4_df
)

# Fit logistic regression model
logit_model <- svyglm(
  formula = gout_status ~ betel_nut,
  design = table4_survey,
  family = binomial(link = "logit")
)

# Display model summary
summary(logit_model)

table_regression <- tbl_regression(logit_model, 
                                   exponentiate = TRUE,
                                   show_single_row = betel_nut,
                                   label = list(
                                     betel_nut ~ "Betel Nut Use"
                                   ),
                                   pvalue_fun= ~ style_pvalue(.x, digits = 3)
                                   ) %>% 
  as_gt() %>%
  gt::tab_header("Table 4: Logistic Regression - Betel Nut Use & Gout") %>% 
  gt::tab_options(
    table.font.size = "small",
    data_row.padding = gt::px(1)
  )

```

Here is what table 4 looks like. 

```{r table4out, echo=TRUE, results='asis'}
print(table_regression)
```

### Validating BRFSS weights and results 
```{r valid, echo=TRUE, message=FALSE}
# Validating BRFSS weights and parameters (test variable & example == alcohol_consumption prevalence)

################## Unweighted Data Analysis ########################
# Calculate the total number of responses
total_responses <- length(guam_data$alcohol)

# Calculate the total number of cases
alcohol_frequency <- table(guam_data$alcohol)

# Calculate the percentages
percentages <- prop.table(alcohol_frequency) * 100

#Display the frequency table and percentages
test_result <- data.frame(Response = names(alcohol_frequency),Frequency = as.numeric(alcohol_frequency), Percentage = percentages)

```


Here is what the <b> unweighted data </b> look like. 


```{r valid1, echo=TRUE, results='asis'}
knitr::kable(test_result, format = "markdown")
```


```{r valid2, echo=TRUE, message=FALSE}
################## Weighted Data Analysis ########################

# Create the survey design
test_survey <- svydesign(
  ids = ~v4,
  strata = ~v3,
  weights = ~v75,
  data = guam_data
)

# Calculate weighted total number of responses
total_responses <- sum(weights(test_survey))

# Calculate the weighted frequency table
weighted_alcohol_frequency <- svytable(~alcohol, design = test_survey)

# Calculate the weighted percentages
weighted_percentages <- prop.table(weighted_alcohol_frequency) * 100

# Display the weighted frequency table and percentages
test_weighted_result <- data.frame(Response = names(weighted_alcohol_frequency),
                              Frequency = as.numeric(weighted_alcohol_frequency),
                              Percentage = weighted_percentages)
```


Here is what the <b> weighted data </b> look like. 


```{r valid3, echo=TRUE, results='asis'}
knitr::kable(test_weighted_result, format = "markdown")
```

Sweet!! The weighted percents and unweighted n's match what's in the CDC website Prevalence tool!
Check Guam 2020 alcohol consumption 30 days - https://www.cdc.gov/brfss/brfssprevalence/

:D

For `gtsummary` documentation please see: https://www.danieldsjoberg.com/gtsummary/index.html 